---
title: "Final Exam Template"
author: "kgteller"
date: "2023-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1 Probabilities Form Random Variables

For binomially distributed random variable with n=35 and p=0.43 find P(X\>22)

```{r}

prob <- 1 - pbinom(18, size = 25, prob = 0.63)
prob
```

For binomially distributed random variable with n=25 and p=0.43 find P(X=22)

```{r}
prob1 <- 1 - dbinom(22, size = 25, prob = 0.43)
prob1
```

For Poisson random variable with mean 6  find P(X=2)

```{r}
prob2 <- dpois(5, 3)

prob2
```

For Poisson random variable with mean 6  find $P(X \leq 2)$

```{r}
prob3 <- ppois(5, 3)

prob3
```

For normal random variable with mean 3 and standard deviation 0.75 find $P(1 \leq X < 2.1)$.

```{r}
prob4 <- pnorm(2.8, mean = 2, sd = .5) - pnorm(1, mean = 2, sd = .5)
prob4
```

For normal random variable with mean 3 and standard deviation 0.75 find $P(X=2)$.

```{r}
prob5 <- dnorm(2, mean = 2, sd = .5)
prob5
```

For exponential random variable with mean 2, find $P(1.1 < X \leq 2.3)$

```{r}
prob5 <- pexp(2.5, rate = 1/2) - pexp(1, rate = 1/2)
prob5
```

## Part 2 Cancer an Well Water Data

1\. Determine how many patients were diagnosed with cancer in each county.

```{r}
library("readxl")
cancer <- read_excel("cancer_and_well_water.xlsx")
cancer
count <- table(cancer$County)
count

```

2\. Create cross-tabulation table of county and stage of cancer patients were diagnosed with

```{r}
cross <- table(cancer$County, cancer$Stage)
cross

745
444

```

3\. Find descriptive statistics of age for all patients.

```{r}
age <- as.numeric(cancer$AgeDx)
sum <- summary(age)
sum
```

4\. Find descriptive statistics of patients' age by stage of cancer they were diagnosed with.

```{r}
cross <- table(cancer$AgeDx, cancer$Stage)
cross

summary(cross)
```

5\. Construct 95% confidence interval for the average age of cancer patients.

```{r}
age <- as.numeric(cancer$AgeDx)
t.test(age, conf.level = 0.95)

```

6\. Test a hypothesis whether average age of all cancer patients exceeds 65 at the time of diagnosis.

```{r}
t.test(age, mu = 65, alternative = "greater")
```

7\. Test a hypothesis to determine if average age of patients diagnosed at an early stage is different from average age of patients diagnosed at an advanced stage.

```{r}

a <- as.numeric(cancer$AgeDx[cancer$Stage == "Early"])
b <- as.numeric(cancer$AgeDx[cancer$Stage == "Advanced"])
t.test(a, b)
```

## Part 3 Mortgage Data and Regression

1\. Create a scatter plot of mortgage rates.

```{r}
mort <- read_excel("mortgage_data.xlsx")
mort
```

2.  Find regression equation modeling 30-year rate using 15-year rate as predictor.

```{r}
model <- lm(mort$`30 year`~mort$`15 year`)
model
summary(model)
```
